Linear stability analysis for \textit{monami} in a submerged sea grass bed

### 2015-06-19 Paper_JFM/grass.tex
Final finishing touches

\definecolor{tableShade}{gray}{0.8}
\title{\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
% \author{L. Mahadevan}
% \affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

\begin{abstract}
The onset of \monami ~-- the synchronous waving of sea grass beds driven by a steady flow -- is modeled as a linear instability of the flow. Our model treats the drag exerted by the grass in establishing the steady flow profile, and in damping out perturbations to it. This damping leads to a finite threshold flow for the instability, which agrees with experimental observations. This role of vegetation drag differentiates our mechanism from the previous hypothesis that the Kelvin-Helmholtz instability underlies \monami.
\end{abstract}

\maketitle
\section{Introduction}
Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with fluid flows.  
The hydrodynamic processes resulting from this behavior influence many environmental processes such as sedimentation, transportation of dissolved oxygen, plant growth, and biomass production  \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}. 
One such response of the submerged grass beds to steady currents is the formation of coherent large amplitude oscillations, known as \monami ~\cite{AckermanOkubo93}.  
In this letter, we provide a hydrodynamic mechanism for the onset of these coherent oscillations.

Current explanations of \monami ~invoke the existence of a free shear layer at the top of the grass bed (henceforth called grass top) due to vegetation drag  \cite{Ikeda96,Ghisal02,Raupach96}. 
Its instability, through a mechanism similar to the Kelvin-Helmholtz instability, is thought to lead to coherent eddies over the grass bed.
The grass responds to these eddies by deforming, which manifests itself as large amplitude synchronous oscillations.
Basic application of this model have been used to predict the observed frequency of \monami~and to understand transport in the seagrass bed \cite{Nepf00,Ghisal02,Nepf04,Okamoto12}.

However, several aspects of this explanation remain unsatisfactory: 
(i)   The velocity profile of the free shear layer is assumed \textit{ad hoc} to be piecewise linear \cite{Delangre06} or hyperbolic tangent \cite{Ghisal02,Raupach96}. 
(ii)  The role of drag in damping the coherent perturbations to the shear profile is sometimes ignored \cite{Raupach96}. 
(iii) No existing theory explains the threshold flow speed, observed in the lab \cite{Ghisal02} and the field \cite{Grizzle96}, below which \monami ~is not observed.
(iv)  The thickness of the free shear layer is in many cases comparable to the unvegetated water depth, and therefore inconsistent with the definition of the free shear layer.

@@

We find two modes of instability labeled Mode 1 and Mode 2.  
We find that Mode 1 resembles many characteristics of Kelvin-Helmholtz instability but with the inclusion of the vegetation drag. 
Mode 2 is not related to the Kelvin-Helmholtz instability mechanism and arises purely due to the interaction of an inviscid layer of fluid with submerged rigid.
The presence of vegetation drag leads to a finite threshold flow condition for onset of the instability for both these modes.

\section{Mathematical model}
Although \monami ~is manifest in the motion of the grass, the drag exerted by the grass bed on the flow is central to the hypothesized instability. 
The instability and the resulting flow structures persist in lab experiments even when flexible grass mimics are replaced by rigid dowels \cite{Ghisal02,Nepf06}. 
Therefore, to develop the essential mathematical model, we assume the grass blades to be rigid, on average oriented vertically.
We show that the drag of the grass results in an instability mode different from the Kelvin-Helmholtz instability.
\begin{figure}
\centerline{\includegraphics[scale=1]{Grass_Base_Nepf_shear}}
% \includegraphics[width=15cm]{Grass_Base_Nepf_shear}
\caption{
Schematic setup and comparison of our steady flow profile with that from the experiments in ref. \cite{Nepf04} (Case I from Table 1) %  with 1250 plants/m$^2$, plant height = 13.7$\pm 0.2$ cm and blade width of 0.64 cm)
 and its approximation with $U_0=7.28$ cm/s and $\delta = 5.02$ cm in our model. The grass extends up to $y=\hg$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into a parabolic profile in the unvegetated region, a uniform profile deep within the vegetation, and a boundary layer of thickness $\delta$ near the grass top. 
The dependence of the boundary layer thickness (estimated as $|U/U_y|$ at $y=\hg$ from the numerical solution of \eqref{base_equ}) on the vegetation density parameter $\ReyNdg$ is shown in the inset.
}
\label{basicflow}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[scale = 0.95]{new_graph_R_vs_delta}\includegraphics[scale = 0.95]{new_graph_K_vs_delta}
% \includegraphics[width=12cm]{Critical_Re_vs_delta_noshear} \\
% \vspace{-6mm} \hspace{-5mm}
% \includegraphics[width=12cm]{K_vs_shear_width_noshear}
\end{center}
\caption{
Critical Reynolds number (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density parametrized by the boundary layer thickness. 
Parameters from experiments reported in ref. \cite{Ghisal02} to exhibit or suppress synchronous waving are also included in the top panel. 
Inset compares experimental observations of the experimentally measured dominant frequency $f_o$ (in Hz) with the predictions $f_p=\text{Im}(\sigma)$ from the solution of ~\eqref{Orr-somerfield}. 
The experimental data in the inset is obtained from refs. \cite{Ghisal02} (Ghisalberti) and \cite{Vivoni98} (Vivoni). 
In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed.
}
\label{Re_vs_delta}
\end{figure}

The drag exerted by the vegetation, assumed dense, is modeled by a continuous body force $\mathbf{f}$.
This drag enters the fluid mass and momentum balance equations as 
\begin{equation}
\grad\cdot\bu = 0,\hspace{3mm} \rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad p+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}
\label{navier-stokes}
\end{equation}
where $\rho$, $\bu$, and $p$ are the fluid density, velocity, and pressure respectively, $\mu$ the dynamic eddy viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow based on the scale of the grass blade is $O(10^2-10^3)$; therefore, neglecting the skin friction, we model it as the form drag on the vegetation, $\mathbf{f}=N_g C_N \rho u |u| d\bx$ \cite{Nepf99,Nepf00,Nepf04}, where  $N_g$ is the blade number density per unit horizontal area, $C_{N}$ the drag coefficient, and $d$ the average blade width projected perpendicular to the flow. 
In the interest of simplicity, we model the turbulence using an eddy viscosity; see \S\ref{subsec:turbulencemodel} for a detailed discussion. 
In the field, both $C_N$, $N_g$ and $\mu$ vary with position, but we do not expect these variations to be central to the instability mechanism, and therefore take them to be constants. 
Based on previous experiments \cite{Vivoni98,Nepf00}, we take $C_N = 1$.

@@

\section{Linear stability analysis}
We first calculate the fully developed steady solution $\bu = U(y)\boldsymbol{\hat{x}}$ of ~\eqref{navier-stokes} driven by constant pressure gradient $dP/dx$, satisfying
\begin{equation}
 -\frac{dP}{dx}+\mu U''(y) +S(y) \rho C_N d N_gU |U|=0
\label{base_equ}
\end{equation}
where $S(y)=1$ for $0<y<\hg$ and $S(y)=0$ for $\hg< y< 2H$. 
Eq. \eqref{base_equ} is solved subject to no shear at the boundaries, i.e., $U'(0) = U'(2H) = 0$.
The former arises for dense vegetation because the shear stress exerted by the bottom surface is expected to be negligible compared to the vegetation drag~\cite{Nepf00}, whereas the latter models the free interface. 
A comparison of the steady flow profile from the solution of ~\eqref{base_equ} with experimental measurements is shown in Fig.~\ref{basicflow}.
The profile $U(y)$ has three distinct regions.
Within vegetation, it is approximately uniform with $ U(y) \approx U_g = \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$, arising from balancing the drag with pressure gradient. 
Outside the vegetation, the velocity has a simple parabolic profile. % due to the balance between viscous forces and the pressure gradient. 
At the grass top, continuity of shear stresses results in a boundary layer of thickness $\delta$. 
Denoting $\ubl$ to be the velocity scale in the boundary layer, and $U_0 = {(dP/dx)~H^2}/{\mu}$ the velocity scale in the unvegetated region, the balance between viscous forces and vegetation drag implies $(\mu \ubl/\delta^2 \sim \rho C_N d N_g \ubl^2)$, and the continuity of shear stress across the grass top implies $(\ubl/\delta \sim U_0/H)$.
Solving for $\delta$ and $\ubl$ yields $\delta/H = \ubl/U_0=(\ReyNdg)^{-1/3}$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, and $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow. 
A numerical estimate of $\delta$ ($U/U_y$ at $y=\hg$) is compared with this prediction in Fig. \ref{basicflow} (inset).
Because this boundary layer develops independent of the influence of the boundaries, we identify it to be analogous to the free shear layer \cite{Ghisal02,Nepf04} in the previous explanation of \monami.
This dependence of $\delta$ on $N_g$, verified in Fig.~\ref{basicflow}, gives us a way to systematically investigate the effect of the shear layer thickness on the instability mechanism.
The figure also shows that the asymptotic regime of a thin boundary layer is expected to hold for $\ReyNdg \gtrsim 100$. In this notation, $U_g/U_0 = (\Rey \Ndg)^{-1/2}$ (used for deriving \eqref{eqn:mode2asymp}).

Next we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in ~\eqref{navier-stokes} and expand to linear order to investigate the evolution of small perturbations $(\tilde{u}, \tilde{v})$, which obey
\begin{equation}
\begin{split}
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0,
\end{split} \nonumber
\end{equation}
where the tilde are dropped.
These equations are non-dimensionalized using half channel height $H$, velocity $U_0$, and time $H/U_0$, leading to three non-dimensional parameters, \textit{viz.} $\Rey$, $\Ndg$, and the vegetation submergence ratio $\hg/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the vegetation density and help elucidate the instability mechanism. 
Using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a solution of 
the form $\left(u,v,\psi \right)= \left(\hat u(y), \hat v(y), \phi(y) \right)e^{ikx+\sigma t}$ to obtain a modified Orr-Sommerfeld equation \cite{Drazin81,Chen97,Chu91} 
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+\Rey \Ndg D\left(2 S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$. 
The growth rate $\sigma$ for a given wave number $k$ appears as an eigenvalue that allows a non-trivial solution $\phi$ of  \eqref{Orr-somerfield}.
We solve \eqref{Orr-somerfield} numerically for $\sigma$ and $\phi$.

\section{Results - unstable modes and critical parameters}
A threshold in $\Rey$, above which the flow is unstable (Re$(\sigma)>0$) for at least one $k$, emerges from the solution of ~\eqref{Orr-somerfield}. 
The dependence of this threshold $\Rey$, and the corresponding marginally stable wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown in Fig.~\ref{Re_vs_delta}, and is found to compare well with experimental observations \cite{Ghisal02}.
The threshold $\Rey$ increases with the vegetation density, indicating a competition between the destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
A similar conclusion was presented for an analogous problem (flow around an emergent (i.e., $\hg>2H$) sea grass patch), but by assuming $U(y)$ to be a tanh-profile, and neglecting the viscous term \cite{White07}.
Previous calculations for terrestrial grass either exclude the vegetation drag in their models \cite{Raupach96}, or assume the mean velocity profile \textit{ad hoc} \cite{Raupach96,Delangre06}.
A threshold flow condition is not reported previously either for terrestrial or submerged marine meadows.

The frequency (Im$(\sigma)$) of the fastest growing mode agrees well with observed frequencies -- frequency of \monami, maxima in the velocity spectra, and frequency of vortex passage in lab scale experiments \cite{Ghisal02} -- for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, as shown in the inset in Fig.~\ref{Re_vs_delta}. 
The experimentally observed \monami ~wavelengths are not available for comparison.

\begin{table*}
% \footnotesize
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& Equation \eqref{eqn:Rayleigh}		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  no grass included & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& yes, near $y=\hg$			& no, spans water column
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two unstable modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the instability mechanism, we consider the dependence of the fastest growing wavenumber on $\delta$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical $\delta$ discontinuously jumps and remains $O(1)$ (see Fig.~\ref{Re_vs_delta}). 
To aid in explaining this behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in Fig.~\ref{K_Re_sigma_set3}. 
The smallest $\Rey$ on the neutral curve (Re$(\sigma)=0$) sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two; we refer to the region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

>>>>

\definecolor{tableShade}{gray}{0.8}
% \title{\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed}
\title{Linear stability analysis for \textit{monami} in a submerged sea grass bed}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
% \author{L. Mahadevan}
% \affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

\begin{abstract}
The onset of \monami ~-- the synchronous waving of sea grass beds driven by a steady flow -- is modeled as a linear instability of the flow. Our model treats the drag exerted by the grass in establishing the steady flow profile, and in damping out perturbations to it. We find two distinct modes of instability, which we label Mode 1 and Mode 2. Mode 1 mechanism is closely related to Kelvin-Helmholtz instability modified to account for vegetation drag, whereas Mode 2 is unrelated to Kelvin-Helmholtz and arises from an interaction of the flow in the vegetated and unvegetated layers. The vegetation damping leads to a finite threshold flow for both these modes. The model predictions for the threshold flow criteria for onset and the resulting oscillation frequency compare well with experimental observations. The inclusion of vegetation drag differentiates our mechanism from the previous analyses of \monami.
\end{abstract}

\maketitle
\section{Introduction}
Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with fluid flows.  
The hydrodynamic processes resulting from this behavior influence many environmental processes such as sedimentation, transportation of dissolved oxygen, plant growth, and biomass production  \citep{Fonseca87,Grizzle96,Nepf99,Nepf2012}. 
One such response of the submerged grass beds to steady currents is the formation of coherent large amplitude oscillations, known as \monami ~\citep{AckermanOkubo93}.  
In this letter, we provide a hydrodynamic mechanism for the onset of these coherent oscillations.

Current explanations of \monami~\citep{Ikeda96,Ghisal02,Raupach96} invoke the existence of a shear layer at the top of the grass bed (henceforth called grass top) due to vegetation drag. 
Its instability, through a mechanism similar to the Kelvin-Helmholtz (KH) instability, is thought to lead to coherent eddies over the grass bed.
The grass responds to these eddies by deforming, which manifests itself as large amplitude synchronous oscillations.
Basic application of this model have been used to predict the observed frequency of \monami~and to understand transport in the seagrass bed \citep{Nepf00,Ghisal02,Nepf04,Okamoto12}.

However, several aspects of this explanation remain unsatisfactory. 
The instability is modeled using the inviscid Rayleigh equation and the role of drag in damping the coherent perturbations to the shear profile is ignored \citep{Raupach96}. 
It is so despite the shear layer itself being produced as a result of the vegetation drag.
Furthermore, the shear layer is assumed to be not influenced by the top and bottom boundary of the domain, although the experimentally measured thickness of the shear layer is in many cases comparable to the unvegetated water depth.
The velocity profile of the free shear layer is assumed \textit{ad hoc} to be piecewise linear \citep{Delangre06} or hyperbolic tangent \citep{Ghisal02,Raupach96}. 
While these profiles qualitatively capture the features of the profile, the dependence of the shear layer thickness on the parameters of the problem remains unexplained.
And finally, no existing theory explains the threshold flow speed, observed in the lab \citep{Ghisal02} and the field \citep{Grizzle96}, below which \monami ~is not observed.

@@

We find two modes of instability labeled Mode 1 and Mode 2.  
We find that Mode 1 resembles many characteristics of KH instability but is modified by the presence of the vegetation drag. 
We also find another mode of instability, Mode 2, unrelated to the KH instability mechanism, arising purely due to the interaction of an inviscid layer of fluid with submerged rigid vegetation.
The presence of vegetation drag leads to a finite threshold flow condition for onset of the instability for both these modes.

\section{Mathematical model}
Although \monami ~is manifest in the motion of the grass, the drag exerted by the grass bed on the flow is central to the hypothesized instability. 
The instability and the resulting flow structures persist in lab experiments even when flexible grass mimics are replaced by rigid dowels \cite{Ghisal02,Nepf06}. 
Therefore, to develop the essential mathematical model, we assume the grass blades to be rigid, on average oriented vertically.
We show that the drag of the grass results in an instability mode different from the KH instability.
\begin{figure}
\centerline{\includegraphics[scale=1]{Grass_Base_Nepf_shear}}
% \includegraphics[width=15cm]{Grass_Base_Nepf_shear}
\caption{
Schematic setup and comparison of our steady flow profile with that from the experiments in ref. \cite{Nepf04} (Case I from Table 1) %  with 1250 plants/m$^2$, plant height = 13.7$\pm 0.2$ cm and blade width of 0.64 cm)
 and its approximation with $U_0=7.28$ cm/s and $\delta = 5.02$ cm in our model. The grass extends up to $y=\hg$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into a parabolic profile in the unvegetated region, a uniform profile deep within the vegetation, and a boundary layer of thickness $\delta$ near the grass top. 
The dependence of the boundary layer thickness (estimated as $|U/U_y|$ at $y=\hg$ from the numerical solution of \eqref{base_equ}) on the vegetation density parameter $\ReyNdg$ is shown in the inset.
}
\label{basicflow}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[scale = 0.95]{new_graph_R_vs_delta}\includegraphics[scale = 0.95]{new_graph_K_vs_delta}
% \includegraphics[width=12cm]{Critical_Re_vs_delta_noshear} \\
% \vspace{-6mm} \hspace{-5mm}
% \includegraphics[width=12cm]{K_vs_shear_width_noshear}
\end{center}
\caption{
Critical Reynolds number (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density parametrized by the boundary layer thickness. 
Parameters from experiments reported in ref. \cite{Ghisal02} to exhibit or suppress synchronous waving are also included in the top panel. 
Inset compares experimental observations of the experimentally measured dominant frequency $f_o$ (in Hz) with the predictions $f_p=\text{Im}(\sigma)$ from the solution of ~\eqref{Orr-somerfield}. 
The experimental data in the inset is obtained from refs. \cite{Ghisal02} (Ghisalberti) and \cite{Vivoni98} (Vivoni). 
In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed.
}
\label{Re_vs_delta}
\end{figure}

The drag exerted by the vegetation, assumed dense, is modeled by a continuous body force $\mathbf{f}$.
This drag enters the fluid mass and momentum balance equations as 
\begin{equation}
\grad\cdot\bu = 0,\hspace{3mm} \rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad p+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}
\label{navier-stokes}
\end{equation}
where $\rho$, $\bu$, and $p$ are the fluid density, velocity, and pressure respectively, $\mu$ the dynamic eddy viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow based on the scale of the grass blade is $O(10^2-10^3)$; therefore, neglecting the skin friction, we model it as the form drag on the vegetation, $\mathbf{f}=N_g C_N \rho u |u| d\bx$ \cite{Nepf99,Nepf00,Nepf04}, where  $N_g$ is the blade number density per unit horizontal area, $C_{N}$ the drag coefficient, and $d$ the average blade width projected perpendicular to the flow. 
In the interest of simplicity, we model the turbulence using an eddy viscosity; see \S\ref{subsec:turbulencemodel} for a detailed discussion. 
In the field, both $C_N$, $N_g$ and $\mu$ vary with position, but we do not expect these variations to be central to the instability mechanism, and therefore take them to be constants. 
% Based on previous experiments \cite{Vivoni98,Nepf00}, we take $C_N = 1$.
% Note that only the product $C_N N_g d$ appears together in our model.

@@

\section{Linear stability analysis}
We first calculate the fully developed steady solution $\bu = U(y)\boldsymbol{\hat{x}}$ of ~\eqref{navier-stokes} driven by constant pressure gradient $dP/dx$, satisfying
\begin{equation}
 -\frac{dP}{dx}+\mu U''(y) +S(y) \rho C_N d N_gU |U|=0
\label{base_equ}
\end{equation}
where $S(y)=1$ for $0<y<\hg$ and $S(y)=0$ for $\hg< y< 2H$. 
Eq. \eqref{base_equ} is solved subject to no shear at the boundaries, i.e., $U'(0) = U'(2H) = 0$.
The former arises for dense vegetation because the shear stress exerted by the bottom surface is expected to be negligible compared to the vegetation drag~\citep{Nepf00}, whereas the latter models the free interface. 
A comparison of the steady flow profile from the solution of ~\eqref{base_equ} with experimental measurements is shown in Fig.~\ref{basicflow}.
The profile $U(y)$ has three distinct regions.
Within vegetation, it is approximately uniform with $ U(y) \approx U_g = \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$, arising from balancing the drag with pressure gradient. 
Outside the vegetation, the velocity has a simple parabolic profile. % due to the balance between viscous forces and the pressure gradient. 
At the grass top, continuity of shear stresses results in a boundary layer of thickness $\delta$. 
Denoting $\ubl$ to be the velocity scale in the boundary layer, and $U_0 = {(dP/dx)~H^2}/{\mu}$ the velocity scale in the unvegetated region, the balance between viscous forces and vegetation drag implies $(\mu \ubl/\delta^2 \sim \rho C_N d N_g \ubl^2)$, and the continuity of shear stress across the grass top implies $(\ubl/\delta \sim U_0/H)$.
Solving for $\delta$ and $\ubl$ yields $\delta/H = \ubl/U_0=(\ReyNdg)^{-1/3}$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, and $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow. 
A numerical estimate of $\delta$ ($U/U_y$ at $y=\hg$) is compared with this prediction in Fig. \ref{basicflow} (inset).
We identify the boundary layer to be analogous to the shear layer \citep{Ghisal02,Nepf04} in the previous explanation of \monami.
This dependence of $\delta$ on $N_g$ gives us a way to systematically investigate the effect of the shear layer thickness on the instability mechanism.
Fig. \ref{basicflow} also shows that the asymptotic regime of a thin boundary layer is expected to hold for $\ReyNdg \gtrsim 100$. 
In this notation, $U_g/U_0 = (\Rey \Ndg)^{-1/2}$ (used for deriving \eqref{eqn:mode2asymp}).

Next we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in ~\eqref{navier-stokes} and expand to linear order to investigate the evolution of small perturbations $(\tilde{u}, \tilde{v})$, which obey
\begin{equation}
\begin{split}
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0,
\end{split} \nonumber
\end{equation}
where the tilde are dropped.
These equations are non-dimensionalized using half channel height $H$, velocity $U_0$, and time $H/U_0$, leading to three non-dimensional parameters, \textit{viz.} $\Rey$, $\Ndg$, and the vegetation submergence ratio $\hg/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the vegetation density and help elucidate the instability mechanism. 
Using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a solution of 
the form $\left(u,v,\psi \right)= \left(\hat u(y), \hat v(y), \phi(y) \right)e^{ikx+\sigma t}$ to obtain a modified Orr-Sommerfeld equation \citep{Drazin81,Chen97,Chu91} 
\begin{equation}
\begin{split}
\Rey^{-1}\left(D^2 -k^{2} \right)^2\phi &= \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi + \Ndg D\left(2 S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$. 
The growth rate $\sigma$ for a given wave number $k$ appears as an eigenvalue that allows a non-trivial solution $\phi$ of  \eqref{Orr-somerfield}.
We solve \eqref{Orr-somerfield} numerically for $\sigma$ and $\phi$.

\section{Results - unstable modes and critical parameters}
A threshold in $\Rey$, above which the flow is unstable (Re$(\sigma)>0$) for at least one $k$, emerges from the solution of ~\eqref{Orr-somerfield}. 
The dependence of this threshold $\Rey$, and the corresponding marginally stable wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown in Fig.~\ref{Re_vs_delta}, and is found to compare well with experimental observations \citep{Ghisal02}.
The threshold $\Rey$ increases with the vegetation density, indicating a competition between the destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
A similar conclusion was presented for an analogous problem (flow around an emergent (i.e., $\hg>2H$) sea grass patch), but by assuming $U(y)$ to be a tanh-profile, and neglecting the viscous term \citep{White07}.
Previous calculations for terrestrial grass either exclude the vegetation drag in their models \cite{Raupach96}, or assume the mean velocity profile \textit{ad hoc} \citep{Raupach96,Delangre06}.
A threshold flow condition is not reported previously either for terrestrial or submerged marine meadows.

The frequency (Im$(\sigma)$) of the fastest growing mode agrees well with observed frequencies -- frequency of \monami, maxima in the velocity spectra, and frequency of vortex passage in lab scale experiments \citep{Ghisal02} -- for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, as shown in the inset in Fig.~\ref{Re_vs_delta}. 
The experimentally observed \monami ~wavelengths are not available for comparison.

\begin{table*}
% \footnotesize
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{l|c|c|c}
			& KH 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& Equation \eqref{eqn:Rayleigh}		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  no grass included & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& yes, near $y=\hg$			& no, spans water column
 \end{tabular}
 \caption{Comparison between KH instability and the two unstable modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the instability mechanism, we consider the dependence of the fastest growing wavenumber on $\delta$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical $\delta$ discontinuously jumps and remains $O(1)$ (see Fig.~\ref{Re_vs_delta}). 
To aid in explaining this behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in Fig.~\ref{K_Re_sigma_set3}. 
The smallest $\Rey$ on the neutral curve (Re$(\sigma)=0$) sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two; we refer to the region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

\section{Discussion -- comparison of unstable modes with Kelvin Helmholtz}
The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ distinguishes it from the KH instability. 
\subsection{Mode 1}
We numerically observe that critical Reynolds number above which perturbation become unstable scales as  $\Rey \sim (H/\delta)^2$ (or $\Rey \propto {\Ndg}^{2}$). 
Our calculations also show that Mode 1 asymptotically localizes to the boundary layer near the grass top, and exhibits highest growth for a perturbation of  $k \sim H/\delta$ (see Fig.~\ref{Re_vs_delta}). 
The behavior of this critical Reynolds number can be understood by considering the limit $\Rey \gg 1$ and $\Ndg \gg 1$.
Estimating the size of various terms of ~\eqref{Orr-somerfield} within the boundary layer in this limit help us understand the behavior of the critical $\Rey$. 
Using $D\sim H/\delta$, $\sigma \sim O(1)$, and $U=\ubl \sim \delta/H$ in the boundary layer; the magnitude of the advection term is $\Rey (H/\delta)^2$  (or $\Rey^{5/3} \Ndg^{2/3}$), and the viscous as well as vegetation drag term are $(\delta/H)^{-4}$ (or $(\Rey \Ndg)^{4/3})$. 
The advection term, viscous term and vegetation drag terms balances when $\Rey \sim (H/\delta)^2$ (or $\Rey \sim {\Ndg}^{2}$).
